New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2023-01-11 paper/paper.tex
compiled version of G changes

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approach
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many tour developments, which are
summarized in \citet{lee2021}.

@@

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods is illustrated
to compare and contrast boundaries constructed by different classifiers.
The software section describes a mathematica package that is used for
the application, and describes the interactive environment that would be
desirable within R as new technology becomes available. The paper is
accompanied by an appendix with more details and adjustments to the
manual controls, and three Mathematica notebooks that can be used to
reproduce the application.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables contributing to a \(d\) dimensional projection. The initial
ingredients are an orthonormal basis (\(A_{p\times d}\)) defining the
projection of the data, and a variable id (\(m \in \{1, ..., p\}\))
specifying which coefficient will be changed. A method to update the
values of the component (\(m^{th}\) row of \(A_{p\times d}\)) of the
controlled variable \(V_m\) is then needed.

@@

One might implement such an interactive interface via R Shiny, but in
particular the tracking of the mouse position in small increments might
pose a challenge. Here we found the dynamic graphics interface from
Mathematica to be especially useful. In the following we first describe
the relevant Mathematica functionalities and how they were used in the
implementation, followed by a brief sketch of how the new approaches are
implemented in the \texttt{tourr} package. There is hope that better
interactivity will be available for R soon, which might allow for
implementation of the techniques developed in Mathematica.

@@

Mathematica provides much utility and versatility, such as inbuilt data
visualization, data manipulation and analysis, dynamic functionality,
and symbolic and numeric computation. Most of this inbuilt functionality
is user-friendly and described in the Mathematica documentation;
typically, numerous examples are provided within the documentation, and
some possible issues are outlined there.

Importantly for our work, it is relatively simple to create dynamic
objects via the inbuilt commands \texttt{Manipulate} or
\texttt{DynamicModule}. Control objects, such as sliders, locator panes,
and input fields, can then be used on dynamic variables, and when there
is a change in the dynamic variable the dynamic objects which contain
that variable will be updated. This is the essential ingredient for this
implementation of the manual tour.

The most relevant inbuilt Mathematica function for our purpose is the
\texttt{LocatorPane}. This creates a region on the screen where the
position of the mouse is captured and then converted to input that
updates the graphics functions, enabling the manual navigation of the
tour.

>>>>

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications,
\textcolor{blue}{methods which work for arbitrary dimensions have more general utility}.
A good approach is to show projections from an arbitrary dimensional
space to create dynamic data visualizations called \emph{tours}. Tours
involve views of high-dimensional (\(p\)) data with low-dimensional
(\(d\)) projections. In his original paper on the grand tour,
\citet{As85} provided several algorithms for tour paths that could
theoretically show the viewer the data \emph{from all sides}. Prior to
Asimov's work, there were numerous preparatory developments including
\citet{tukey}'s PRIM-9. PRIM-9 had user-controlled rotations on
coordinate axes, allowing one to manually tour through low-dimensional
projections. (A video illustrating the capabilities is available through
\textcolor{blue}{the} video library of \citet{ASA22}.) Steering through
all possible projections is impossible, unlike Asimov's tours which
allows one to quickly see many, many different projections. After Asimov
there have been many tour developments, which are summarized in
\citet{lee2021}.

@@

This paper explains the new manual controls for projection and slice
tours. The next section describes the new algorithm for manual control,
for both projections and slices. The use of these methods is illustrated
to compare and contrast boundaries constructed by different classifiers.
The software section describes a Mathematica \citet{Mathematica} package
that is used for the application, and describes the interactive
environment that would be desirable within R \citet{rref} as new
technology becomes available. The paper is accompanied by an appendix
with more details and adjustments to the manual controls, and three
Mathematica notebooks that can be used to reproduce the application.

@@

A manual tour allows the user to alter the coefficients of one (or more)
variables contributing to a \(d\) dimensional projection. The initial
ingredients are an orthonormal basis (\(A_{p\times d}\)) defining the
projection of the data, and a variable id (\(m \in \{1, ..., p\}\))
specifying which coefficient will be
\textcolor{blue}{changed: the $m^{th}$ row of $A_{p\times d}$ which we denote by $V_m$.}
A method to update the values of the component of the controlled
variable \(V_m\) is then needed.

@@

One might implement such an interactive interface via R Shiny
\citet{shinym}, but in particular the tracking of the mouse position in
small increments might pose a challenge. Here we found the dynamic
graphics interface from Mathematica to be especially useful. In the
following we first describe the relevant Mathematica functionalities and
how they were used in the implementation, followed by a brief sketch of
how the new approaches are implemented in the \texttt{tourr} package.
There is hope that better interactivity will be available for R soon,
which might allow for implementation of the techniques developed in
Mathematica.

@@

\textcolor{blue}{The Mathematica} implementation relies on the inbuilt
Mathematica function \texttt{LocatorPane}. This function creates a
region on the screen where the position of the mouse is captured and
then converted to input that updates the graphics functions, enabling
the manual navigation of the tour. We then use the graphics function
\texttt{ListPlot} embedded in dynamic objects via \texttt{DynamicModule}
to produce the display. Control objects, such as sliders, locator panes,
and input fields, are used on dynamic variables, and when there is a
change in the dynamic variable the dynamic objects which contain that
variable will be updated.

Di will explain what part and why is difficult in R, also check
grammar\}
